Multiset-trie data structure

### 2022-12-09 tex/referee-data/response.txt

2. Has the statistical analysis been performed appropriately and rigorously?
Reviewer #1: N/A

3. Have the authors made all data underlying the findings in their manuscript fully available?

Reviewer #1: No

NAŠ ODZIV NA ZGORNJE:
We published the data that was used to preform experimets from the paper, and refered to those accordingly.

? (to ask Mikita, both iztok and matjaž
Matjaž takes care of publishing data)

@@

Reviewer #1: Yes

5. Review Comments to the Author

Reviewer #1:
The motivation is not clear, the authors claim that this is an important problem in object-relational databases, information retrieval, data mining and expert systems. However, for example, when authors claim to be a relevant problem in information retrieval, they provide two citations to complete books of information retrieval, without any indication of section or chapter. Please, please be more specific. A good practice is to show the specific problems (examples) of the application.

NAŠ ODZIV NA ZGORNJE:
We followed the advice of referee, and replaced references with the more specific ones, as requested.
This can be found in the second pharagraph of the introduction (Section 1), where we overview the related areas of computer science.
Notice that in those presented areas the set containment operations play an essential role.

---

@@

It would be nice if the authors provide running examples of their algorithms, they would be easier to understand.

NAŠ ODZIV NA ZGORNJE:
We acknowledge the remark of the reviewer, however we feel that this paper is already substentilly long, and providing an additional figure for each operation might not give the desirable outcome.

Please notice that we did include an example of the underlying multiset-trie data structure.

---

@@

All figures are of very poor quality and can hardly be read.

NAŠ ODZIV NA ZGORNJE:
All figures presenting various plots are generated by matplotlib python library, and exported in vector format.
We believe this to be industry-standard and kindly ask for detailed explanation on how to improve their quality.
%Forthermore we believe that those figures were carefully designed

@@

NAŠ ODZIV NA ZGORNJE:

In multisets ordering is not important.
We use a specific ordering of the multiset elements to speed up the set containment queries.
This specific ordering is used when we construct the multiset-trie, so the
ordering is a part of the data structure.

We have updated the introduction (Section 1) to:
1) emphasise that the ordering is not important in multisets, and
2) to depict the utilization of the selected ordering for multiset containment operations.

---

The experiments are not convincing for the same reason, as the whole paper obviates the ordering, whereas the inverted index uses a B-tree to search for the inverted list of each element in the multiset.
The experiments must consider that the multisets are not ordered nor added, that is, an input query could be {1, 2, 2, 1, 3, 3, 1, 2, 2}

@@

NAŠ ODZIV NA ZGORNJE:

We kindly point out that the specification of all our algorithms is designed so that the input queries are already in the so-called Multiset type (see Section 3 for more details).
This type assumes an order to the input elemens, as well as precomputed multiplicity of each of them.
We may have two distinct orderings of multisets.
The first is the application ordering where particular element can be at many places in the multiset.
The second ordering is the ordering of the symbols that is used for implementation of multiset-trie.
In particular this implies that the preprocessing step from the above comment is not covered by our complexity analysis.
We acknowledge that different applications may require the choice of our definitions to be modified.

----
The values of n, the maximum multiplicity of each element of the multiset, are small, most experiments (as shown in Table 1) are 3, and one with 10. Experiments with larger values of n are needed, and, as I will argue next, authors must include the size of the resulting data structures. The scalability must be studied, and experiments showing how the size of n affects the query times and the size of the data structure must be included.

NAŠ ODZIV:
We acknowledge that our description of the multiset-trie data structure is not optimized, as we instead strive for a clean represeantation of a new idea.
We also point out that the absence of the compression makes the multiset-trie a perfectly height-balanced tree, which simplifies the analysis.

That said, there are various ways on how to optimize this data structure further.
For example, the multiplicity of an element can be implemented using a small, customly designed, hash table (instead of a simple array as in the paper).

In fact, we added this and some other possible optimiziations to the conclusion of our paper.

-----

@@

The idea of minimality of the data structure size stated in line 808 page 27, for me, is wrong. A simple analysis reveals that the space complexity O(|M|) is a very coarse bound, and it is not minimal. First, the size of the data structure is better bounded by the number of nodes, which is analyzed in Section 4.1.1. Second, the space efficiency is not optimal because you are using a complex method to determine the label of a node (the symbol a node contains). That is, to know the label of a node, it is necessary to know the position of the pointer that points to the node of interest, within the \sigma pointers of the parent node. Then each node needs an array of \sigma entries, even though only some of them are filled, so you are wasting many empty entries. In addition, since the label of a node is given by a pointer, unless you use a modern tree (that exists, but you have not mentioned that) I assume that you require 32 bits to represent the pointer (label) of a node. Instead, if you explicitly represent the symbol of each node, you only need log \sigma bits.

NAŠ ODZIV NA ZGORNJE:

We agree with the reviewer, and we dropped the comment about the space minimality of the data
structure in the conclusions (Section 7), accordingly.
The minimality of the data structure is not mentioned in any other part of the paper.

@@

Finally, the intersection algorithm used in your experiments is not described. There are different intersection algorithms, please refer to, for example :
J. Shane Culpepper and Alistair Moffat. 2011. Efficient set intersection for inverted indexing. ACM Trans. Inf. Syst. 29, 1, Article 1 (December 2010), 25 pages. https://doi.org/10.1145/1877766.1877767

That with the better behaviour in your case must be selected.

NAŠ ODZIV NA ZGORNJE:

We understand the comment of reviewer. However, one of the authors of this paper recently went to work in industry.
As he was in charge of the development of inverted index implementation, it seems that a new implementation is hardly possible at the moment.
That said, we agree that in the current form the Experiment 5 does not fairly compare both data structures.
Accordingly, we have omitted the empirical comparison of the multiset-trie with the inverted file from the paper.

-----

@@

Related work is poor, especially the explanation of the set-trie. Although Section 6.1 sketches the differences between multiset-trie and set-trie, more explanations are needed. First, the differences in the data structures and then, the differences in operations must be clarified.

NAŠ ODZIV NA ZGORNJE:

We have added a more detailed presentation of the set-trie in the related work.
Most notable changes can be found in the second paragraph of Section 6.1.
In the current form, also differences in operations between the two data structures are, in our opinion, exposed better (see third paragraph of Section 6.1).

### 2022-12-09 tex/tex/s_math-Theorems.tex
MDPI template
threw everything into MDPI template.
created parsed file for math section (uses {Theorem} instead of {theorem})

\section{Mathematical analysis of the structure} \label{c:analysis}
%
In this chapter we present theoretical results of time and space complexity of
the multiset-trie data structure. In the following Section~\ref{s:timecomplexity} 
we discuss the running time complexity of the presented algorithms. First, in 
Section~\ref{ss:mathmodel}, we present our mathematical model that we use 
to describe the distribution of multisets in the multiset-trie and input data. 
Using probabilistic approach and tools from a Galton-Watson process we 
measure the expected cardinality of the multiset-trie in Theorem~\ref{thm:exp_nodes}. 
Further, we derive the expected cardinality of the searched subtree of the 
multiset-trie parametrized by an input multiset in Corollary~\ref{cor:exp_nodes_param}.

In Section~\ref{ss:getall} we discuss the running time complexity of the functions 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets}. We observe that the 
complexity of functions is exponential. Moreover, the worst case running time 
complexity is the same for both functions and its upper bound is the cardinality of 
the multiset-trie.

The remaining "existence" functions are discussed in the Section~\ref{ss:exists}. 
We observe that out of scope of our mathematical model unlike in functions 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets} the mapping $\phi$ has an 
impact on performance of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence}. In particular, the frequency analysis of the symbols 
from $\Sigma$ in input data determines such a $\phi$ that gives a boost in performance.

We find that the performance of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} in the worst case scenario is also exponential and 
does not depend on the outcome of the functions. We give a quite precise upper 
bound for the worst case running time complexity, which appears to be the same 
for both functions. However, it must be stressed that for the positive outcome an 
exponential behavior holds only on specific cases, such as presence of the emptyset 
in the multiset-trie.

Finalizing the mathematical analysis, we present the study of space complexity 
of the multiset-trie in the Section~\ref{s:spacecomplexity}. We show that the space 
used for the storage is asymptotically equal to the size of the input data.

\subsection{Time complexity of the algorithms}\label{s:timecomplexity}
The performance of the functions will be measured by the number of 
visited nodes in a multiset-trie during execution of a particular query by the 
functions \textsc{search}, \textsc{delete}, \textsc{submsetExistence}, 
\textsc{supermsetExistence}, \textsc{getAllSubmsets}, \textsc{getAllSupermsets} 
and the procedure \textsc{insert}.
%

By the design of the multiset-trie, it is easy to see that the functions \textsc{search},
\textsc{delete} and the procedure \textsc{insert} have complexity of $O(\sigma).$
Because $\sigma$ is defined when the structure is initialized and does not depend
on the user input afterwards, the asymptotic complexity of the functions \textsc{search}, 
\textsc{delete} and the procedure \textsc{insert} is $O(1).$ Nonetheless, in the general 
case the complexity is $O(\sigma).$

In what follows, we focus on analysis of the more involved functions:
\textsc{submsetExistence}, \textsc{supermsetExistence}, \textsc{getAllSubmsets}
and \textsc{getAllSupermsets}.

\subsubsection{Mathematical model} \label{ss:mathmodel}
We start with the basics of our mathematical model. Let $\Sigma$ be an alphabet of
cardinality $\sigma,$ such that $\Sigma = \{ 1,2, \ldots, \sigma \}.$ Define $N$
to be the set of all possible multisets that can be inserted in a
multiset-trie. Let $n$ be the maximal degree of a node in a multiset-trie.
Then the maximal multiplicity of an element in a multiset is equal to $n-1.$
Thus, the number of multisets in a complete multiset-trie is $ |N| = n^{\sigma}.$
%
Let $M$ be a collection of multisets inserted into multiset-trie $\mathcal{M}.$
All the multisets in $M$ are constructed from the alphabet $\Sigma$ according
to the parameters $\sigma$ and $n$. Hence, any multiset $m\in M,$ 
has at most $\sigma$ distinct elements that are members of $\Sigma$ and 
every distinct element in $m$ has multiplicity strictly less than $n.$
%
Because a multiset does not distinguish different orderings, it is assumed, for
simplicity that all elements are ordered in an ascending order. A multiset $m$ is
represented as $\{1^{k_1},2^{k_2},\ldots, \sigma^{k_\sigma}\},$ where $e^{k_e}$
represents an element $e\in\Sigma$ with multiplicity $k_e.$
%

Denote the nodes of multiset-trie on all levels but on $\sigma + 1$ as \emph{internal}
and nodes on leaf level as \emph{leaf} nodes.
%
Observe that every internal non-root node has a degree at least~1. Indeed an
insertion of a multiset requires a construction of a path of length~$\sigma + 1,$
meaning that if an internal node exists in a multiset-trie it must have a degree
at least~1. It also follows that the height of a multiset-trie is always~$\sigma +1$
as soon as at least one multiset is inserted into the data structure.

Our model assumes that all the inserted multisets are chosen with the same probability,
meaning that for some $p\in (0,1)$ the following holds:
\[
P(m\in M) = p, \quad \forall~m\in N.
\]
%
Let $\xi_1, \xi_2, \ldots, \xi_{\sigma+1}$ be random variables such that $\xi_i$
represents the number of nodes in a multiset-trie on $i$-th level. For every node $j$ 
on $i$-th level we assign a random variable $\xi_{ij}$ to be the number of its children, 
such that $j\in[1,\xi_i].$ Then for every $i\in[1,\sigma]$ the following holds:
%
\begin{equation}\label{eq:sum_recursive}
\xi_{i+1} = \sum_{j=1}^{\xi_i} \xi_{ij},
\end{equation}
%
where $\xi_1 = 1.$
%
It is easy to see that the variable $\xi_{i+1}$ can have values in the interval
$[\xi_i,n^{i}]$ and the value of the variable $\xi_{ij}$ is within the interval $[1,n].$
Without conditioning on the existence of any node in multiset-trie, 
it is easy to describe the probability of existence of any individual node.

\begin{lemma}\label{l:prob-node-existence}
Any potential node on a fixed level $i,$ where $i\in \{ 1,2,\ldots, \sigma +1 \}$ exists, %independently of others,
with probability
\begin{equation}
p_i=1-(1-p)^{n^{\sigma + 1 -i}}.
\end{equation}
\end{lemma}
\begin{proof}
Let $v$ be an arbitrary node in a multiset-trie on an arbitrary level $i.$ Consider
the subtree with the root $v$ and call it $v$-subtree. Since the height of the
multiset-trie is $\sigma + 1$ we can calculate the height of the $v$-subtree.
Taking in account that the root node has height 1, the height of the $v$-subtree is
%
\[
h_v = \sigma + 1 - i.
\]
%
A node in a multiset-trie exists if at least one node exists on the leaf level of its
subtree, i.e. a node on the level $\sigma + 1$ that belongs to $v$-subtree. The possible
number of nodes on the leaf level of $v$-subtree can be easily calculated knowing its height.
It is equal to
%
\[
n^{\sigma + 1 - i}.
\]
%
A node at level $\sigma+1$ exists with probability $p,$ where $p = P(m\in M).$
Thus, the probability that there are no nodes on leaf level in $v$-subtree is
%
\[
(1-p)^{n^{\sigma +1 - i}}.
\]
%
The claim follows by taking the complement probability of the above result.

@@

\begin{lemma}\label{l:prob-children}
Suppose that a node $v$ exists at level $1\leq i\leq\sigma$.
Then the number of its children $\xi_{iv}$ is modeled by a zero-truncated binomially
distributed random variable on parameters $n$ and $p_{i+1}$. In particular,
the probability of node $v$ having $k$ children equals to
\begin{equation}\label{eq:pdf}
P(\xi_{iv} = k) = \frac{\binom{n}{k} (1-p_{i+1})^{n-k}}{1-(1-p_{i+1})^n}
\end{equation}
and the corresponding probability generating function equals to
\begin{equation}\label{eq:generating_func}
G_i(z) = \frac{(1+p_{i+1}(z-1))^n - (1-p_{i+1})^n}{1-(1-p_{i+1})^n}.
\end{equation}
\end{lemma}
\begin{proof}
In order to prove the lemma, we have to show that
$\xi_{iv}\sim\mathcal{B}_0(n, p_{i+1}).$
Consider an arbitrary node $v$ on level $1\leq i\leq\sigma.$ According to the
definition of the multiset-trie a node exists at level $i$ if and only if
it has at least one child. Note that this is not true for the nodes on the leaf
level $\sigma + 1.$ Implies, a node on level $i$ can have
$k\in\{ 1,2,\ldots, n\}$ children. Let $X_0, X_1, \ldots, X_{n-1}$ be random
variables, they are defined as follows:
\[
X_k = \begin{cases}
0 & \textrm{child $k$ of node $v$ does not exist} \\
1 & \textrm{child $k$ of node $v$ exists} \\
\end{cases}
\]
As it was shown in previous Lemma~\ref{l:prob-children}, the distribution of 
$X_k$ is $X_k\sim Bernoulli(p_{i+1}).$ Since our model assumes that all the 
multisets in $M$ are chosen uniformly at random, the variables 
$X_k,X_l$ are independent for $k\neq l.$ But in our case the node $v$ can not 
have 0 children, so the sum $\sum_{k=1}^n X_k$ has a zero-truncated binomial 
distribution:
%
\[
\sum_{k=1}^n X_k \sim\mathcal{B}_0(n,p_{i+1})
\]
%
which completes the proof.

\end{proof}
%
Knowing the probability density and probability generating functions of $\xi_{ij}$ 
from Lemma~\ref{l:prob-children}, we now can estimate the number of nodes in 
a randomly generated multiset-trie as follows:
%
\begin{equation}\label{eq:num_nodes}
\mathbb{E}( | \mathcal{M} | ) = \mathbb{E}\left[ \sum_{i=1}^{\sigma+1} \xi_i \right].
\end{equation}
%

@@

In order to evaluate~(\cref{eq:num_nodes}) we will use some of the tools from 
a Galton-Watson process, see Gardiner~\cite{gardiner1985stochastic} for 
an introduction.
Using the equations~(\cref{eq:sum_recursive}) and~(\cref{eq:generating_func}) we can 
derive the probability generating function for the random variable $\xi_{i+1}$ as 
\begin{equation}\label{eq:g_rec}
G_{\xi_{i+1}}(z) = G_{\xi_i}(G_i(z)).
\end{equation}
Since there is always precisely one node at the root-level, we have $P(\xi_1 = 1) = 1.$ 
Hence, the probability generating function for the random variable $\xi_1$ is 
\begin{equation}\label{eq:g_xi_one}
G_{\xi_1}(z) = z^1 = z
\end{equation}
which is the initial condition for the recursive equation~(\cref{eq:g_rec}).

\begin{proposition}\label{prop:exp_rec}
The expectation of the random variable $\xi_{i+1}$ can be expressed 
as follows.
\[
\mathbb{E}(\xi_{i+1}) = \mathbb{E}(\xi_i)\mathbb{E}(\mathcal{B}_0(n,p_{i+1}))
\]
for $1\leq i\leq\sigma.$
\end{proposition}
\begin{proof}
Using the following property of probability generating function
\begin{equation}\label{eq:prop_exp-pgf}
G_X'(1^-) = \mathbb{E}(X)
\end{equation}
the expectation for the random variable $\xi_{i+1}$ can be derived in terms of 
the equation~(\cref{eq:g_rec}).
\begin{eqnarray}\label{eq:exp}
\mathbb{E}(\xi_{i+1}) &=& G_{\xi_{i+1}}'(1^-) \nonumber \\
&=& G_{\xi_i}'(G_i(1^-))G_i'(1^-).
\end{eqnarray}
According to~(\cref{eq:pdf}) and~(\cref{eq:generating_func}) the value of $G_i(z)$ at 
$1$ is $1$ and the value of its derivative at $1$ is $\mathbb{E}(\mathcal{B}_0(n,p_{i+1})).$ 
Substituting the values of $G_i(1^-)$ and $G_i'(1^-),$ and applying the 
property~(\cref{eq:prop_exp-pgf}) we complete the proof.

@@

\begin{theorem}\label{thm:exp_level}
Let $\mathcal{M}$ be a multiset-trie defined with parameters $n,$ $\sigma,$ and denote the number 
of nodes on every level $i$ by a random variable $\xi_i.$ Furthermore, let all multisets appear in 
$\mathcal{M}$ with equal probability $p\in (0,1).$ Then the expected number 
of nodes on every level of $\mathcal{M},$ i.e. $\mathbb{E}(\xi_i)$ is defined as 
\begin{equation}\label{eq:nodes_level}
\mathbb{E}(\xi_{i}) = n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}.
\end{equation}
\end{theorem}
\begin{proof}
According to~(\cref{eq:g_xi_one}) the expected number of nodes on the first level is 1. \\
Using $\mathbb{E}(\xi_1) = 1$ and the result from Proposition~\ref{prop:exp_rec}
we get
\begin{eqnarray*}
\mathbb{E}(\xi_{i}) &=& \prod_{j=2}^{i} \frac{n p_j}{1-(1-p_j)^n}
= \prod_{j=2}^{i} n \frac{1-(1-p)^{n^{\sigma +1-j}}}{1-(1-p)^{n^{\sigma + 2 -j}}} \\
&=& n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}
\end{eqnarray*}

>>>>

\section{Mathematical analysis of the structure} \label{c:analysis}
%
In this chapter, we present theoretical results of time and space complexity of
the multiset-trie data structure. In the following Section~\ref{s:timecomplexity} 
we discuss the running time complexity of the presented algorithms. First, in 
Section~\ref{ss:mathmodel}, we present the mathematical model that we use 
to describe the distribution of multisets in the multiset-trie and input data. 
Using a probabilistic approach and tools from a Galton-Watson process, we 
measure the expected cardinality of the multiset-trie in Theorem~\ref{thm:exp_nodes}. 
Further, we derive the expected cardinality of the searched subtree of the 
multiset-trie parametrized by an input multiset in Corollary~\ref{cor:exp_nodes_param}.

In Section~\ref{ss:getall} we discuss the running time complexity of the functions 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets}. We observe that the 
complexity of functions is exponential. Moreover, the worst-case running time 
complexity is the same for both functions, and its upper bound is the cardinality of 
the multiset-trie.

The remaining "existence" functions are discussed in the Section~\ref{ss:exists}. 
We observe that out of the scope of our mathematical model, unlike in functions 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets} the mapping $\phi$ has an 
impact on performance of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence}. In particular, the frequency analysis of the symbols 
from $\Sigma$ in input data determines such a $\phi$ that gives a boost in performance.

We find that the performance of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} in the worst-case scenario is also exponential and 
does not depend on the outcome of the functions. We give a quite precise upper 
bound for the worst-case running time complexity, which appears to be the same 
for both functions. However, it must be stressed that for the positive outcome, an 
exponential behavior holds only on specific cases, such as the presence of the empty set 
in the multiset-trie.

Finalizing the mathematical analysis, we present the study of space complexity 
of the multiset-trie in Section~\ref{s:spacecomplexity}. We show that the space 
used for the storage is asymptotically equal to the size of the input data.

\subsection{Time complexity of the algorithms}\label{s:timecomplexity}
The performance of the functions will be measured by the number of 
visited nodes in a multiset-trie during the execution of a particular query by the 
functions \textsc{search}, \textsc{delete}, \textsc{submsetExistence}, 
\textsc{supermsetExistence}, \textsc{getAllSubmsets}, \textsc{getAllSupermsets} 
and the procedure \textsc{insert}.
%

By the design of the multiset-trie, it is easy to see that the functions \textsc{search},
\textsc{delete} and the procedure \textsc{insert} have complexity of $O(\sigma).$
Because $\sigma$ is defined when the structure is initialized and does not depend
on the user input afterwards, the asymptotic complexity of the functions \textsc{search}, 
\textsc{delete} and the procedure \textsc{insert} is $O(1).$ Nonetheless, in the general 
case, the complexity is $O(\sigma).$

In what follows, we focus on the analysis of the more involved functions:
\textsc{submsetExistence}, \textsc{supermsetExistence}, \textsc{getAllSubmsets}
and \textsc{getAllSupermsets}.

\subsubsection{Mathematical model} \label{ss:mathmodel}
We start with the basics of our mathematical model. Let $\Sigma$ be an alphabet of
cardinality $\sigma,$ such that $\Sigma = \{ 1,2, \ldots, \sigma \}.$ Define $N$
to be the set of all possible multisets that can be inserted in a
multiset-trie. Let $n$ be the maximal degree of a node in a multiset-trie.
Then the maximal multiplicity of an element in a multiset is equal to $n-1.$
Thus, the number of multisets in a complete multiset-trie is $ |N| = n^{\sigma}.$
%
Let $M$ be a collection of multisets inserted into multiset-trie $\mathcal{M}.$
All the multisets in $M$ are constructed from the alphabet $\Sigma$ according
to the parameters $\sigma$ and $n$. Hence, any multiset $m\in M,$ 
has at most $\sigma$ distinct elements that are members of $\Sigma$ and 
every distinct element in $m$ has multiplicity strictly less than $n.$
%
Because a multiset does not distinguish different orderings, it is assumed, for
simplicity, that all elements are ordered in ascending order. A multiset $m$ is
represented as $\{1^{k_1},2^{k_2},\ldots, \sigma^{k_\sigma}\},$ where $e^{k_e}$
represents an element $e\in\Sigma$ with multiplicity $k_e.$
%

Denote the nodes of multiset-trie on all levels but on $\sigma + 1$ as \emph{internal}
and nodes on leaf level as \emph{leaf} nodes.
%
Observe that every internal non-root node has a degree at least~1. Indeed an
insertion of a multiset requires construction of a path of length~$\sigma + 1,$
meaning that if an internal node exists in a multiset-trie, it must have a degree
at least~1. It also follows that the height of a multiset-trie is always~$\sigma +1$
as soon as at least one multiset is inserted into the data structure.

Our model assumes that all the inserted multisets are chosen with the same probability,
meaning that for some $p\in (0,1)$, the following holds:
\[
P(m\in M) = p, \quad \forall~m\in N.
\]
%
Let $\xi_1, \xi_2, \ldots, \xi_{\sigma+1}$ be random variables such that $\xi_i$
represents the number of nodes in a multiset-trie on $i$-th level. For every node $j$ 
on $i$-th level we assign a random variable $\xi_{ij}$ to be the number of its children, 
such that $j\in[1,\xi_i].$ Then for every $i\in[1,\sigma]$ the following holds:
%
\begin{equation}\label{eq:sum_recursive}
\xi_{i+1} = \sum_{j=1}^{\xi_i} \xi_{ij},
\end{equation}
%
where $\xi_1 = 1.$
%
It is easy to see that the variable $\xi_{i+1}$ can have values in the interval
$[\xi_i,n^{i}]$ and the value of the variable $\xi_{ij}$ is within the interval $[1,n].$
Without conditioning on the existence of any node in multiset-trie, 
it is easy to describe the probability of the existence of any individual node.

\begin{Lemma}\label{l:prob-node-existence}
Any potential node on a fixed level $i,$ where $i\in \{ 1,2,\ldots, \sigma +1 \}$ exists, %independently of others,
with probability
\begin{equation}
p_i=1-(1-p)^{n^{\sigma + 1 -i}}.
\end{equation}
\end{Lemma}
\begin{proof}
Let $v$ be an arbitrary node in a multiset-trie on an arbitrary level $i.$ Consider
the subtree with the root $v$ and call it $v$-subtree. Since the height of the
multiset-trie is $\sigma + 1$ we can calculate the height of the $v$-subtree.
Taking into account that the root node has height 1, the height of the $v$-subtree is
%
\[
h_v = \sigma + 1 - i.
\]
%
A node in a multiset-trie exists if at least one node exists on the leaf level of its
subtree, i.e. a node on the level $\sigma + 1$ that belongs to $v$-subtree. The possible
number of nodes on the leaf level of $v$-subtree can be easily calculated knowing its height.
It is equal to
%
\[
n^{\sigma + 1 - i}.
\]
%
A node at level $\sigma+1$ exists with probability $p,$ where $p = P(m\in M).$
Thus, the probability that there are no nodes on leaf level in $v$-subtree is
%
\[
(1-p)^{n^{\sigma +1 - i}}.
\]
%
The claim follows by taking the complement probability of the above result.

@@

\begin{Lemma}\label{l:prob-children}
Suppose that a node $v$ exists at level $1\leq i\leq\sigma$.
Then the number of its children $\xi_{iv}$ is modeled by a zero-truncated binomially
distributed random variable on parameters $n$ and $p_{i+1}$. In particular,
the probability of node $v$ having $k$ children equals to
\begin{equation}\label{eq:pdf}
P(\xi_{iv} = k) = \frac{\binom{n}{k} (1-p_{i+1})^{n-k}}{1-(1-p_{i+1})^n}
\end{equation}
and the corresponding probability generating function equals to
\begin{equation}\label{eq:generating_func}
G_i(z) = \frac{(1+p_{i+1}(z-1))^n - (1-p_{i+1})^n}{1-(1-p_{i+1})^n}.
\end{equation}
\end{Lemma}
\begin{proof}
In order to prove the lemma, we have to show that
$\xi_{iv}\sim\mathcal{B}_0(n, p_{i+1}).$
Consider an arbitrary node $v$ on level $1\leq i\leq\sigma.$ According to the
definition of the multiset-trie, a node exists at level $i$ if and only if
it has at least one child. Note that this is not true for the nodes on the leaf
level $\sigma + 1.$ Implies, a node on level $i$ can have
$k\in\{ 1,2,\ldots, n\}$ children. Let $X_0, X_1, \ldots, X_{n-1}$ be random
variables, they are defined as follows:
\[
X_k = \begin{cases}
0 & \textrm{child $k$ of node $v$ does not exist} \\
1 & \textrm{child $k$ of node $v$ exists} \\
\end{cases}
\]
As it was shown in previous Lemma~\ref{l:prob-children}, the distribution of 
$X_k$ is $X_k\sim Bernoulli(p_{i+1}).$ Since our model assumes that all the 
multisets in $M$ are chosen uniformly at random, the variables 
$X_k,X_l$ are independent for $k\neq l.$ But in our case the node $v$ can not 
have 0 children, so the sum $\sum_{k=1}^n X_k$ has a zero-truncated binomial 
distribution:
%
\[
\sum_{k=1}^n X_k \sim\mathcal{B}_0(n,p_{i+1})
\]
%
which completes the proof.

\end{proof}
%
Knowing the probability density and probability generating functions of $\xi_{ij}$ 
from Lemma~\ref{l:prob-children}, we can now estimate the number of nodes in 
a randomly generated multiset-trie as follows:
%
\begin{equation}\label{eq:num_nodes}
\mathbb{E}( | \mathcal{M} | ) = \mathbb{E}\left[ \sum_{i=1}^{\sigma+1} \xi_i \right].
\end{equation}
%

@@

In order to evaluate~(\ref{eq:num_nodes}) we will use some of the tools from 
a Galton-Watson process, see Gardiner~\cite{gardiner1985stochastic} for 
an introduction.
Using the equations~(\ref{eq:sum_recursive}) and~(\ref{eq:generating_func}) we can 
derive the probability generating function for the random variable $\xi_{i+1}$ as 
\begin{equation}\label{eq:g_rec}
G_{\xi_{i+1}}(z) = G_{\xi_i}(G_i(z)).
\end{equation}
Since there is always precisely one node at the root level, we have $P(\xi_1 = 1) = 1.$ 
Hence, the probability generating function for the random variable $\xi_1$ is 
\begin{equation}\label{eq:g_xi_one}
G_{\xi_1}(z) = z^1 = z
\end{equation}
which is the initial condition for the recursive equation~(\ref{eq:g_rec}).

\begin{Proposition}\label{prop:exp_rec}
The expectation of the random variable $\xi_{i+1}$ can be expressed 
as follows.
\[
\mathbb{E}(\xi_{i+1}) = \mathbb{E}(\xi_i)\mathbb{E}(\mathcal{B}_0(n,p_{i+1}))
\]
for $1\leq i\leq\sigma.$
\end{Proposition}
\begin{proof}
Using the following property of probability generating function
\begin{equation}\label{eq:prop_exp-pgf}
G_X'(1^-) = \mathbb{E}(X)
\end{equation}
the expectation for the random variable $\xi_{i+1}$ can be derived in terms of 
the equation~(\ref{eq:g_rec}).
\begin{eqnarray}\label{eq:exp}
\mathbb{E}(\xi_{i+1}) &=& G_{\xi_{i+1}}'(1^-) \nonumber \\
&=& G_{\xi_i}'(G_i(1^-))G_i'(1^-).
\end{eqnarray}
According to~(\ref{eq:pdf}) and~(\ref{eq:generating_func}) the value of $G_i(z)$ at 
$1$ is $1$ and the value of its derivative at $1$ is $\mathbb{E}(\mathcal{B}_0(n,p_{i+1})).$ 
Substituting the values of $G_i(1^-)$ and $G_i'(1^-),$ and applying the 
property~(\ref{eq:prop_exp-pgf}) we complete the proof.

@@

\begin{Theorem}\label{thm:exp_level}
Let $\mathcal{M}$ be a multiset-trie defined with parameters $n,$ $\sigma,$ and denote the number 
of nodes on every level $i$ by a random variable $\xi_i.$ Furthermore, let all multisets appear in 
$\mathcal{M}$ with equal probability $p\in (0,1).$ Then the expected number 
of nodes on every level of $\mathcal{M},$ i.e. $\mathbb{E}(\xi_i)$ is defined as 
\begin{equation}\label{eq:nodes_level}
\mathbb{E}(\xi_{i}) = n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}.
\end{equation}
\end{Theorem}
\begin{proof}
According to~(\ref{eq:g_xi_one}) the expected number of nodes on the first level is 1. \\
Using $\mathbb{E}(\xi_1) = 1$ and the result from Proposition~\ref{prop:exp_rec}
we get
\begin{align*}
\mathbb{E}(\xi_{i}) &= \prod_{j=2}^{i} \frac{n p_j}{1-(1-p_j)^n}
= \prod_{j=2}^{i} n \frac{1-(1-p)^{n^{\sigma +1-j}}}{1-(1-p)^{n^{\sigma + 2 -j}}} \\
&= n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}
\end{align*}
\end{proof}

\begin{Theorem}\label{thm:exp_nodes}
The expected cardinality of a multiset-trie defined on parameters $n,$ $\sigma$ 
and $p$ can be computed as 
\begin{equation}
\mathbb{E}( | \mathcal{M} | ) = \sum_{i=1}^{\sigma+1} n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}},
\end{equation}
where $r = (1-p)^n,$ so $r\in(0,1).$
\end{Theorem}
\begin{proof}
Using the results obtained from Theorem~\ref{thm:exp_level} we compute 
\begin{align*}
\mathbb{E}( | \mathcal{M} | ) &= \mathbb{E}\left[ \sum_{i=1}^{\sigma+1} \xi_i \right] \\
&= \sum_{i=1}^{\sigma + 1} n^{i-1} \frac{1-(1-p)^{n^{\sigma + 1 -i}}}{1-(1-p)^{n^{\sigma}}}
\end{align*}
\end{proof}

With the expected number of nodes in a multiset-trie $\mathcal{M}$ obtained from 
Theorem~\ref{thm:exp_nodes}, we can now generalize the result for a subtree 
in $\mathcal{M}$ parametrized by an input multiset $m.$ The subtrees that we 
are interested in are the ones that contain all the sub-multisets or all the 
super-multisets of $m.$ In order to calculate the expected cardinality of such 
subtrees, we need the following definition.

\begin{Definition}\label{def:params}
Let $m = \{ 1^{k_1}, 2^{k_2}, \ldots, \sigma^{k_\sigma} \},$ where 
$e^{k_e}$ is an element $e$ with multiplicity $k_e.$ Let $M_1, M_2$ be 
the subsets of the set $M,$ such that $M_1 = \{ x\in M : x\subseteq m \}$ and 
$M_2 = \{ x\in M : x\supseteq m \}.$
Define $\alpha_i$ and $\beta_i$ as follows 
\begin{equation*}
\alpha_i = \begin{cases}
1, & i=0 \\
\prod_{j=1}^i (k_j+1), & 1\leq i\leq\sigma \\
\end{cases}
\end{equation*} 
and 
\begin{equation*}
\beta_i = \begin{cases}
1, & i=0 \\
\prod_{j=1}^i (n-k_j-1), & 1\leq i\leq\sigma \\
\end{cases}.
\end{equation*}
\end{Definition}
